---
title: "STAT 346 Final Project"
subtitle: "Can Money Buy Happiness?: Understanding the Differential Between a Country’s GDP Rank and Happiness Rank"
author: "Edith Edwards-Mizel, Varya Kluev, Greta Laesch"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message = F, warning = F}
library(pander)   
library(car)  
library(ggplot2)
library(gridExtra)
library(datasets)
library(knitr)
library(GGally)
library(dplyr)
library(readr)
library(reshape2)
library(kableExtra)
```


## Adding and Cleaning Data
```{r, message=FALSE}
# Explanatory Variables
census_data <- read_csv("~/Desktop/export.csv", show_col_types = FALSE)
who_data <- read_csv("~/Desktop/Annex 2-4-Table 1.csv", show_col_types = FALSE)
gdp_data <- read_csv("~/Desktop/GDP_data.csv", show_col_types = FALSE)
wh_data <- read_csv("~/Desktop/WHD.csv", show_col_types = FALSE)
who_data <- who_data[-(1:4),]
colnames(census_data)[4] ="country"
colnames(who_data)[1] ="country"
colnames(who_data)[2] ="Clean_energy_access_prop"
colnames(who_data)[3] ="Harmful_air_mean_conc"
colnames(who_data)[4] ="Mortality_homicide_rate"
colnames(who_data)[5] ="Adult_obesity_rate"
colnames(who_data)[6] ="Tobacco_use_rate"
colnames(who_data)[7] ="Expenditure_health_perc"
colnames(who_data)[8] ="Mortality_suicide_rate"
colnames(who_data)[9] ="Alcohol_consuption_liters"
df_merge <- merge(census_data,who_data, by = "country")
df_merge <- df_merge[-c(2:5) ]

# Categorical Variable
hc_data <- read_csv("~/Desktop/UNHC.csv", show_col_types = FALSE)
univ <- hc_data$country
univ_var <- c()
for (i in 1:nrow(df_merge)) {
  ifelse(df_merge[i,1] %in% univ, univ_var[i] <- 1, univ_var[i] <- 0)
}
df_merge["Univ_hc"] <- univ_var
df_merge$Univ_hc <- as.factor(df_merge$Univ_hc)

# Covid Deaths
covid_deaths <- read_csv("~/Desktop/covidData.csv", show_col_types = FALSE)
# -- Cumulative Deaths from 01/03/2020-12/31/2022
newest_covid <- covid_deaths[covid_deaths$Date_reported == "2022-12-31", ]
cv_dummy_df <- data.frame(newest_covid$Country, newest_covid$Cumulative_deaths)
colnames(cv_dummy_df)[1] <- "country"
colnames(cv_dummy_df)[2] <- "tot_deaths"
df_merge <- merge(df_merge, cv_dummy_df, by = "country")

# Corruption
cpi_data <- read_csv("~/Desktop/CPI.csv", show_col_types = FALSE)
# -- Lower Scores == More Corruption
cpi_data <- cpi_data[-(1:2),]
cpi_dummy_df <- data.frame(cpi_data$`Corruption Perceptions Index 2022: Global scores`, cpi_data$...4)
colnames(cpi_dummy_df)[1] <- "country"
colnames(cpi_dummy_df)[2] <- "cpi_score"
df_merge <- merge(df_merge, cpi_dummy_df, by = "country")
df_merge$cpi_score <- as.numeric(df_merge$cpi_score)

# Response Variable
gdp_dummy_df <- data.frame(gdp_data$country, gdp_data$rank)
colnames(gdp_dummy_df)[1] <- "country"
colnames(gdp_dummy_df)[2] <- "GDP_rank"

df_merge <- merge(df_merge, gdp_dummy_df, by = "country")

wh_dummy_df <- data.frame(wh_data$country, wh_data$rank)
colnames(wh_dummy_df)[1] <- "country"
colnames(wh_dummy_df)[2] <- "WH_rank"

df_merge <- merge(df_merge, wh_dummy_df, by = "country")

df_merge$GDP_rank_std <- ((df_merge$GDP_rank) - mean(df_merge$GDP_rank))/sd(df_merge$GDP_rank)
df_merge$WH_rank_std <- ((df_merge$WH_rank) - mean(df_merge$WH_rank))/sd(df_merge$WH_rank)

df_merge["DIF_rank"] <- (df_merge$GDP_rank_std - df_merge$WH_rank_std)

# Cleaning Data
# Make missing data NA
for (i in 1:length(df_merge)) { # columns
  for (j in 1:nrow(df_merge)) { # rows
    ifelse(df_merge[j,i] == "-", df_merge[j,i] <- NA, NA)
  }
}

# Round <1 to 0
for (i in 1:length(df_merge)) { # columns
  for (j in 1:nrow(df_merge)) { # rows
    ifelse(df_merge[j,i] == "<1" || df_merge[j,i] == "<0.1", df_merge[j,i] <- 0, NA)
  }
}

# Convert Characters to Numeric
for (i in (8:15)) {
  df_merge[,i] <- as.numeric(df_merge[,i])
}

# Rename columns with spaces
colnames(df_merge)[2:7] <- c("Population", "rate_nat_increase", "tot_fert_rate", "life_exp_males", "life_exp_females", "death_rate")
attach(df_merge)





#write.csv(df_merge, "/Users/Greta/Desktop/master_df.csv", row.names=TRUE)
```

## Goal: Understanding the Phenomenon in General

### Exploratory Data Analysis  

```{r, warning=FALSE}
# Response Variables
plot(WH_rank_std ~ GDP_rank_std)
mod_1 <- lm(WH_rank_std~GDP_rank_std)
summary(mod_1)
res1 <- plot(residuals(mod_1) ~ fitted.values(mod_1))
res1

# Explanation: We can see that GDP explains some, but not all of variation in the happiness rating
# of a country, so what else is there to explain why some countries are happier than others,
# compared to 

# Census Data Pairs
ggpairs(df_merge[,c(2,3,4,5,6,7,23)])
# WHO Data Pairs
ggpairs(df_merge[,c(8,9,10,11,12,13,14,15,17,18,23)])

```

### Transformations
```{r}
par(mfrow=c(3,5))
plot(DIF_rank~Population)
plot(DIF_rank~rate_nat_increase)
plot(DIF_rank~tot_fert_rate)
plot(DIF_rank~life_exp_males)
plot(DIF_rank~life_exp_females)
plot(DIF_rank~death_rate)
plot(DIF_rank~Clean_energy_access_prop)
plot(DIF_rank~Harmful_air_mean_conc)
plot(DIF_rank~Mortality_homicide_rate)
plot(DIF_rank~Adult_obesity_rate)
plot(DIF_rank~Tobacco_use_rate)
plot(DIF_rank~Expenditure_health_perc)
plot(DIF_rank~Alcohol_consuption_liters)
plot(DIF_rank~Mortality_suicide_rate)
plot(DIF_rank~cpi_score)

# Transform Non-Linear Relationships as Necessary
df_merge$PopulationT <- log(df_merge$Population)
df_merge$tot_fert_rateT <- log(df_merge$tot_fert_rate)
df_merge$Harmful_air_mean_concT <- log(df_merge$Harmful_air_mean_conc)
df_merge$Mortality_homicide_rateT <- log(df_merge$Mortality_homicide_rate)
df_merge$Mortality_suicide_rateT <- log(df_merge$Mortality_suicide_rate)
df_merge$cpi_scoreT <- log(df_merge$cpi_score)
```


### Preliminary Data Analysis: Model Selection Using Automatic Selection Procedure
```{r, warning=FALSE}
df_merge_noNA <- na.omit(df_merge)
# Original Model Before Transformations
myformula <- formula(paste(colnames(df_merge)[23], paste(colnames(df_merge_noNA)[2:18], collapse = "+"), sep = " ~ "))
mod_noT <- lm(myformula, data = df_merge)
summary(mod_noT)
residualPlot(mod_noT)

# Model with Transformed Non-Linear Variables
mod_T <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_males + life_exp_females + death_rate + Clean_energy_access_prop + Harmful_air_mean_concT + Mortality_homicide_rateT + Adult_obesity_rate + Tobacco_use_rate + Expenditure_health_perc + Mortality_suicide_rateT + Alcohol_consuption_liters + Univ_hc + tot_deaths + cpi_score, data = df_merge)
summary(mod_T)
residualPlot(mod_T)

# Model Selection
mod_noNA <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_males + life_exp_females + death_rate + Clean_energy_access_prop + Harmful_air_mean_concT + Mortality_homicide_rateT + Adult_obesity_rate + Tobacco_use_rate + Expenditure_health_perc + Mortality_suicide_rateT + Alcohol_consuption_liters + Univ_hc + tot_deaths + cpi_scoreT, data = df_merge_noNA)
# Summary matches model with missing data
bw_aic2 = step(mod_noNA, direction = "backward")

# AIC-Determined Model
# ----- So different depending on using transformed or not
mod_bwAIC <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_females + death_rate + Mortality_homicide_rateT + Univ_hc, data = df_merge) # Use the full data set
summary(mod_bwAIC)
residualPlot(mod_bwAIC)

# Final Model 
ggpairs(df_merge[,c(24,3,25,6,7,10,16, 23)])
```

### Investigate Groupings by Principal Component Analysis
```{r}
# Non-Transformed Data
pcp = prcomp(df_merge_noNA[, -c(1,16,19,20,21,22,23,24,25,26,27)], center = TRUE, scale = TRUE)
biplot(pcp, cex = 0.5, xlabs = df_merge_noNA$country)

# Transformed Data
pcpT = prcomp(df_merge_noNA[, -c(1,2,4,9,10,14,18,16,19,20,21)], center = TRUE, scale = TRUE)
biplot(pcpT, cex = 0.5, xlabs = df_merge_noNA$country)

# Final Model (Excluding categorical variable)
pcpT = prcomp(df_merge_noNA[, -c(1:7,9:20,21)], center = TRUE, scale = TRUE)
biplot(pcpT, cex = 0.5, xlabs = df_merge_noNA$country)
```  


### Check Multicollinearity 
```{r}
vif(mod_bwAIC)
```

### Investigate the Countries that are High Leverage Points
```{r}
influencePlot(mod_bwAIC, col=c(1,1))

"Low Leverage/High Residuals:"
df_merge[66,1]
"High Leverage/Low Residuals:"
df_merge[90,1]
"High Residual/Low Leverage/High Cooks Distance"
df_merge[124,1]
"High Residual/High Cooks Distance:"
df_merge[15,1]
"High Leverage/High Cooks Distance/Low Residual:"
df_merge[107,1]

# We see know reason to remove these countries, however we will examine them and also view them in light of the models
```

### Investigate Subsets of Countries

As the biplot showed, there was clustering of countries based on their economic 
development status, therefore we are interested in investigating how subsets
of countries divided by their economic development status influences the model.
We subsetted countries into thirds by their GDP rank, then fitted the final
model on each subset.
```{r, warning = FALSE}
df_merge_first <- subset(df_merge, GDP_rank < 74)
df_merge_second <- subset(df_merge, GDP_rank < 138 & GDP_rank > 74)
df_merge_third <- subset(df_merge, GDP_rank > 138)

"First Subset Model"
mod_first <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_females + death_rate + Mortality_homicide_rateT + Univ_hc, data = df_merge_first)
summary(mod_first)
residualPlot(mod_first)

"Second Subset Model"
mod_second <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_females + death_rate + Mortality_homicide_rateT + Univ_hc, data = df_merge_second)
summary(mod_second)
residualPlot(mod_second)

"Third Subset Model"
mod_third <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_females + death_rate + Mortality_homicide_rateT + Univ_hc, data = df_merge_third)
summary(mod_third)
residualPlot(mod_third)

# The model only fulfills assumptions about residuals for the middle subset of our data.
```

### Remodeling By Subset
```{r}
# Subset 1: Model Selection
df_merge_firstnoNA <- na.omit(df_merge_first)
mod_S1 <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_males + life_exp_females + death_rate + Clean_energy_access_prop + Harmful_air_mean_concT + Mortality_homicide_rateT + Adult_obesity_rate + Tobacco_use_rate + Expenditure_health_perc + Mortality_suicide_rateT + Alcohol_consuption_liters + Univ_hc + tot_deaths + cpi_scoreT, data = df_merge_firstnoNA)

bw_aicS1 = step(mod_S1, direction = "backward")
# ---- Final Model: Mortality_suicide_rateT, life_exp_females, life_exp_males,
# Clean_energy_access_prop, cpi_scoreT, Adult_obesity_rate, Univ_hc

```


```{r}
bw_aicS1_mod <- lm(DIF_rank~life_exp_males + life_exp_females + Clean_energy_access_prop + Adult_obesity_rate + Mortality_suicide_rateT + Univ_hc + tot_deaths + cpi_scoreT, data = df_merge_first)
summary(bw_aicS1_mod)
plot(bw_aicS1_mod$residuals ~ bw_aicS1_mod$fitted.values)
vif(bw_aicS1_mod)
```


```{r}

# Subset 2: Model Selection
df_merge_secondnoNA <- na.omit(df_merge_second)
mod_S2 <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_males + life_exp_females + death_rate + Clean_energy_access_prop + Harmful_air_mean_concT + Mortality_homicide_rateT + Adult_obesity_rate + Tobacco_use_rate + Expenditure_health_perc + Mortality_suicide_rateT + Alcohol_consuption_liters + Univ_hc + tot_deaths + cpi_scoreT, data = df_merge_secondnoNA)

bw_aicS2 = step(mod_S2, direction = "backward")
# ---- Final Model: Mortality_suicide_rateT, life_exp_females, death_rate,
# tot_fert_rateT, Clean_energy_access_prop, cpi_scoreT, Mortality_homicide_rateT

```


```{r}
bw_aicS2_mod <- lm(DIF_rank~tot_fert_rateT + life_exp_females + death_rate + Mortality_homicide_rateT + Expenditure_health_perc + Alcohol_consuption_liters + cpi_scoreT, data = df_merge_second)
summary(bw_aicS2_mod)
plot(bw_aicS2_mod$residuals ~ bw_aicS2_mod$fitted.values)
vif(bw_aicS2_mod)
```


```{r}

# Subset 3: Model Selection
df_merge_thirdnoNA <- na.omit(df_merge_third)
mod_S3 <- lm(DIF_rank~PopulationT + rate_nat_increase + tot_fert_rateT + life_exp_males + life_exp_females + death_rate + Clean_energy_access_prop + Harmful_air_mean_concT + Mortality_homicide_rateT + Adult_obesity_rate + Tobacco_use_rate + Expenditure_health_perc + Mortality_suicide_rateT + Alcohol_consuption_liters + Univ_hc + tot_deaths + cpi_scoreT, data = df_merge_thirdnoNA)

bw_aicS3 = step(mod_S3, direction = "backward")
# ---- Final Model: Mortality_suicide_rateT, life_exp_females, life_exp_males, tot_deaths,
# rate_nat_increase, tot_fert_rateT, Adult_obesity_rate, Clean_energy_access_prop
```
```{r}
bw_aicS3_mod <- lm(DIF_rank~rate_nat_increase + tot_fert_rateT + life_exp_males + life_exp_females + Clean_energy_access_prop + Adult_obesity_rate + tot_deaths, data = df_merge_third)
summary(bw_aicS3_mod)
plot(bw_aicS3_mod$residuals ~ bw_aicS3_mod$fitted.values)
vif(bw_aicS3_mod)

df_merge_second

```
```{r}
## Looking at remaining outliers from individual models

influencePlot(bw_aicS1_mod, col=c(1,1))

"High Residual, high cooks distance"
df_merge[122, 1]
df_merge[59, 1]

"High Leverage, low residual, low studentized residual"
df_merge[95, 1]
df_merge[101, 1]

influencePlot(bw_aicS2_mod, col=c(1,1))
"High Residual, high cooks distance"
df_merge[121, 1]
df_merge[15, 1]

"High Leverage, low residual, low studentized residual"
df_merge[37, 1]
df_merge[41, 1]
         
influencePlot(bw_aicS3_mod, col=c(1,1))
"High Residual, high cooks distance, low leverage"
df_merge[127, 1]
df_merge[89, 1]
df_merge[64, 1]

"High Leverage, low residual, low studentized residual"
df_merge[52, 1]
```


### Model Diagnostics
```{r}
par(mfrow=c(2,2))
plot(mod_bwAIC)
```






